Reduced plasticity and variance in physiological rates for terrestrial ectotherm populations under climate change

### 2024-02-05 ms.qmd
Ok, simplified section and now on to "signature of past climate"

One explanation for why terrestrial ectotherms show minimal acclimation capacity may be related to the fact that terrestrial ectotherms were, on average, acclimated for significantly less time than ectotherms from aquatic habitats. However, rates of acclimation have been shown to be faster for many terrestrial groups compared to aquatic organisms [e.g., amphibians and reptiles have higher rates of acclimation than fishes; See @Einum2023-lz]. Faster rates of acclimation would therefore result in the opposite patterns to those that we observed -- in other words, terrestrial species would be more likely to exhibit lower $lnRR_{Q_{10_{acclimation}}}$ compared to $lnRR_{Q_{10_{acute}}}$ when controlling for acclimation time. The weak acclimation responses of terrestrial ectotherms suggest that evolving plasticity  in physiological rates may be hard, possibly because terrestrial ectotherms are more likely to experience more variable and unpredictable thermal environments than aquatic ectotherms (see below). Reductions in physiological rate variance for terrestrial ectotherms may be because of selective mortality at higher temperatures reducing intrapopulation variability or populations being pushed closer to their thermal maxima (or a combination of both). Disentangling these hypotheses will require empirical studies to connect survival measures at different temperatures with thermal physiology across individuals (e.g., $CT_{max}$).

### ***Plasticity and variance in physiologcial rates do not differ between life-stages***
Interestingly, there were no major differences in acclimation or reductions in physiological variance between life-stages across habitat types. Greater plasticity for early life-stages of development is expected to be important for dealing with the effects of climate change because early life-stages are usually particularly vulnerable periods in development [@martin2015age; @stearns1976life]. However, our results highlight that early life stages are not likely to be protected from the impacts of climate change. Evidence across the literature conflicts in terms of the extent to which one life-stage is more or less plastic than another [@carter2020life; @moghadam2019heat]. For example, @moghadam2019heat showed that larvae *Drosophila* were more plastic in their heat hardening responses compared to adults. In contrast, @carter2020life showed greater thermal plasticity in metabolism for adults but little in pupae of *Onthophagus taurus*. Across a diversity of taxa, we show that there were generally similar patterns between early and late life stages -- both in terms of variance changes and the capacity for plasticity. Unsurprisingly, there is much variation in the literature and our findings suggest that patterns will depend on the species and traits in question. However, very few studies measured multiple life-history stages within the same population. As such, we need more studies that explicitly compare, within populations, differences in plasticity among life stages across multiple traits. A better understanding of the microthermal enviroments experienced by different life-stages will also be important in understanding why plasticity (or lack thereof) may differ among life-stages.

>>>>

### ***Plasticity and variance in physiological rates do not differ between life-stages***
Interestingly, there were no major differences in acclimation or reductions in physiological variance between life-stages across habitat types. Greater plasticity for early life-stages of development is expected to be important to increase resilience to the effects of climate change because early life-stages are often particularly vulnerable periods in development [@martin2015age; @stearns1976life]. However, our results highlight that early life stages are not likely to be protected from the impacts of climate change. Evidence across the literature is equivocal with respect to the extent to which one life-stage is more or less plastic than another [@carter2020life; @moghadam2019heat]. For example, @moghadam2019heat showed that larval *Drosophila* were more plastic in their heat hardening responses compared to adults. In contrast, @carter2020life showed greater thermal plasticity in metabolism for adults but little in pupae of *Onthophagus taurus*. Across a diversity of taxa, we show that there were generally similar patterns between early and late life stages -- both in terms of variance changes and the capacity for plasticity. Unsurprisingly, there is much variation in the literature and our findings suggest that patterns will depend on the species and traits in question. However, very few studies measured multiple life-history stages within the same population. In future research, there needs to be more studies that explicitly compare differences in plasticity among life stages across multiple traits within the same populations. A better understanding of the microthermal environments experienced by different life-stages will also be important in understanding why plasticity may or may not differ among life-stages.
